---
layout: default
title: T-test 101
subtitle: What is t-test and why it works
categories: category1
body-class: categoryclass
tags: [R, NHST]
---


T-test is one of the basic tests used in biology (and perhaps most 
others scientific disciplines). However, it is not always interpreted correctly.
Hopefully, this small demo will help understanding what the t-test is and 
what question is answers.

Let's generate a sample.
```{r generate_sample}
set.seed(0)
N <- 5
sample_x <- rnorm(N, mean=0, sd=1)
plot(sample_x, ylim=c(min(0,sample_x),max(0,sample_x)))
abline(h=0, col='red')
mean(sample_x)
```

Now the question is if the mean of the sample is different from some value.
Typically the question is (or can be reduce down to) if the sample mean is 
different from zero. A quantity or statistic how far away the sample mean from
the NULL mean is so-called [t-statistic](https://en.wikipedia.org/wiki/T-statistic).
\\[ t = \frac{\bar{x}-\mu}{\frac{s}{\sqrt{N}}} \\]

Let's compute the t-statitic assuming the NULL mean value = 0.
```{r, sample_t_stat}
mu_null <- 0
sample_t_stat <- (mean(sample_x)-mu_null)/(sd(sample_x)/sqrt(N))
print(sample_t_stat)
```

Hence, our doubt that the mean of the sample is actually not any different from
the NULL value can be posed in a more quantitative way. That is:

*What is the probability of achieving as big or more extreme t-statistic 
for a sample with the mean equal to the NULL mean?*

Let's calculate this probability directly by generating 1 million random 
samples with mean equal to zero.

As a first step we'll generate random samples with mean equal to the NULL mean.
The actual value of the standard deviation at this point does not matter.
The t-statistic is scaled by the standard deviation anyway.
```{r, simulation}
N_simulations <- 1000000
sim_x <- replicate(N_simulations, rnorm(N,mu_null,sd=1))
print(sim_x[,1:5])
```

Calculating t-statistics for 1 million random samples.
```{r, sim_t_stats}
sim_x_mean_diff <- colMeans(sim_x) - mu_null
sim_x_std_err <- apply(sim_x, 2, sd)/sqrt(N)
sim_t_stats <- sim_x_mean_diff/sim_x_std_err 
```

How often the t-statistic from the random samples is more extreme then the 
t-statistic of the tested sample?

First let's make it clear why we are concerned with both higher and lower 
deviations. It is because we do not say specifically if the sample mean is higher
then or lower then the NULL mean. Our sceptical statement is just that it is 
different. Difference can go both ways - higher and lower. Thus we count 
deviations on both tails.
```{r, two_tails}
hist(sim_t_stats, 100)
abline(v=c(-sample_t_stat,+sample_t_stat), col='red')

equal_or_more_extreme <- abs(sim_t_stats) >= abs(sample_t_stat)
calc_p_val <- sum(equal_or_more_extreme)/N_simulations
print(calc_p_val)
```

The actual t-test p-value is:
```{r}
print(t.test(sample_x)$p.value)
```

Very close!
